On-the-fly Algorithms for GKAT Equivalences \thanks{Identify applicable funding agency here. If none, delete this.}

### 2025-01-22 main.tex
grammar fix for Background on Coalgebra and GKAT

\author{\IEEEauthorblockN{Cheng Zhang\textsuperscript{\textsection}}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University College London}\\
London, United Kingdom \\
0000-0002-8197-6181}
\and
\IEEEauthorblockN{Qiancheng Fu}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
qcfu@bu.edu}
\and
\IEEEauthorblockN{Hang Ji}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
email address or ORCID}
\and
\IEEEauthorblockN{Ines Santacruz}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
email address or ORCID}
\and
\IEEEauthorblockN{Marco Gaboardi}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
email address or ORCID}
}

@@

Guarded Kleene Algebra with Tests, or GKAT~\cite{smolka_GuardedKleeneAlgebra_2020}, is a deterministic fragment of Kleene Algebra with Tests. 
The syntax of GKAT over a set of primitive actions \(K\) and a set of primitive tests \(T\) can be defined in two sorts, boolean expressions \(\BExp\) and GKAT expressions \(\Exp\):
\begin{align*}
    a, b, c ∈ \BExp 
        & ≜ 1 ∣ 0 ∣ t ∈ T ∣ b ∧ c ∣ b ∨ c ∣ \overline{b} \\  
    e, f ∈ \Exp 
        & ≜ p ∈ K ∣ b ∈ \BExp ∣ e +_b f ∣ e ; f ∣ e^{(b)} 
\end{align*}
where \(e +_b f\) is the if statement with condition \(b\), \(e;f\) is the sequencing of expression \(e\) and \(f\), and \(e^{(b)}\) is the while loop with body \(e\) and condition \(b\).
We use notation like \(b ≤ c\), \(b ≡ c\), and \(b ≢ c\) for the usual order, equivalence, and inequivalence in Boolean Algebra.
A GKAT expression can be unfolded into a KAT expression in the usual manner~\cite{kozen_KleeneAlgebraTests_1997c}:
\begin{align*}
    e +_b f & ≜ b; e + \overline{b}; f &
    e^{(b)} & ≜ (b; e)^*; \overline{b}.
\end{align*}
Then the semantics of each expression \(⟦e⟧\) can be computed by the semantics of Kleene Algebra with tests~\cite{kozen_KleeneAlgebraTests_1997c}.
An important construct in the semantics is \emph{atoms}, which are conjunctions of all the primitive tests either in its positive or negative form: for \(T ≜ \{t₁, t₂, …, tₙ\}\)
\[\At_T ≜ \{t₁' ∧ t₂' ∧ ⋯ ∧ tₙ' ∣ tᵢ' ∈ \{tᵢ, \overline{tᵢ}\}\}.\]
Follow the conventional notation, we denote an atom using \(α, β\); and we sometimes omit the subscript \(T\) when no confusing can arise.
Alternatively, atoms can also be thought of as truth assignments to each primitive tests, indicating which primitive tests is satisfied in the current program states; and \(α ≤ b\) if and only if the truth assignment represented by \(α\) satisfies \(b\).

For the sake of brevity, we omit the complete definition of GKAT and KAT semantics, we refer the reader to previous works~\cite{smolka_GuardedKleeneAlgebra_2020,schmid_GuardedKleeneAlgebra_2021,kozen_KleeneAlgebraTests_1997c}, which explains these semantics in detail.
Our work avoids direct interaction with the semantics by leveraging prior results in the coalgebraic theory of GKAT, which we will recap below.

@@

Besides nice property with bisimulation, GKAT coalgebra is also deterministic, unlike Kleene coalgebra with tests (KCT)~\cite{kozen_CoalgebraicTheoryKleene_2017}.
For each atom, states in a KCT can accept or reject the atom (but not both), yet states can also non-deterministically transition to multiple different states via the same atom, while executing different actions.
As we will see later, the deterministic behavior of GKAT coalgebra not only enables a more versatile symbolic algorithm than KCT~\cite{pous_SymbolicAlgorithmsLanguage_2015}, but also present challenges. 
Specifically, GKAT coalgebra requires normalization to compute finite trace equivalences~\cite{smolka_GuardedKleeneAlgebra_2020}, where we reroute all the transitions that cannot lead to acceptance immediately into rejection; and states that can never lead to acceptance is called \emph{dead states}.

In previous works, these dead states are detected after the necessary coalgebra is computed and stored.
This approach requires storing all the transition and states of coalgebra in memory, meaning that the algorithm does not short circuit even if a mismatch can be detected in the starting state; for example, when for some \(α ∈ \At\), \(δ_S(s, α) = \reject\) and \(δ_U(u, α) = \accept\).

Our algorithm is lazy in the dead state detection i.e. the dead state are only checked when a mismatch requires it; even then, our dead state detection algorithm will only check the necessary states to compute the liveness of the states causing the mismatch.
This not only avoids unnecessarily checking the liveness of a state, but also enables allows on-the-fly generation of the coalgebra, like using derivatives, where we can remove irrelevant states from memory after it has been explored.

However, to truly understand our on-the-fly algorithms, we will first need to define ``dead states'', and its role in defining the coalgebraic semantic of GKAT.

@@

Traditionally, live and dead states are defined by whether they can reach an accepting state~\cite{smolka_GuardedKleeneAlgebra_2020}.
\begin{definition}[Liveness]\label{def:liveness-of-states}
    A state \(s\) is \emph{accepting} if there exists an atom \(α ∈ \At\) s.t. \(δ(s, α) = \accept\).
    A state \(s₀\) is \emph{live} if there exists an accepting state \(sₙ\) s.t. there is a liveness chain:
    \[s₀ \transvia{α₁ ∣ p₁} s₁ \transvia{α₂ ∣ p₂} s₂ ⋯ \transvia{αₙ ∣ pₙ} sₙ.\]
    A state is \emph{dead} if it is not dead.
\end{definition}
However, besides directly working with this explicit definition, we can also give a more coalgebraic characterization of these properties.
Specifically, by induction on the length of the path from one state to the next, we can show that the principle sub-coalgebra \(⟨s⟩_S\) exactly contains all the reachable state of \(s\) in the GKAT coalgebra \(S\).
Thus we obtain the following coalgebraic characterization of liveness:
\begin{theoremEnd}{theorem}\label{thm:liveness-principle-sub-coalg}
    A state \(s\) is \emph{live} if there exists an accepting state \(s' ∈ ⟨s⟩\); and a state \(s\) is \emph{dead} if there is no accepting state in \(⟨s⟩\).
\end{theoremEnd}
This alternative liveness definition can help us prove important theorems regarding reachability and liveness without explicitly performing induction on the liveness chain. 
We show the following theorems as examples:
\begin{theoremEnd}{lemma}\label{thm:dead-iff-all-reachable-dead}
    A state \(s\) is dead if and only if all elements in \(⟨s⟩\) is dead.
\end{theoremEnd}
\begin{proofEnd}
    \(⟸\) direction is true, because \(s ∈ ⟨s⟩\): if all \(⟨s⟩\) is dead, then \(s\) is dead. 
    \(⟹\) direction can be proven as follows.
    Take any \(s' ∈ ⟨s⟩\), then \(⟨s'⟩ ⊑ ⟨s⟩\) because \(s'\) is the minimal sub-coalgebra that contains \(s'\). 
    Since there is no accepting state in \(⟨s⟩\), thus there cannot be any accepting state in \(⟨s'⟩\), hence \(s'\) is also dead.
\end{proofEnd}

@@

Besides giving us the finite-trace semantics, the normalization operation also satisfy several nice properties.

@@

Second, normalization is an endofunctor in the category of GKAT coalgebra, this result can help us obtain sub-coalgebras of normalized coalgebra, and also connect the finite trace and infinite trace semantics.

>>>>

\author{\IEEEauthorblockN{Cheng Zhang\textsuperscript{\textsection}}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University College London}\\
London, United Kingdom \\
0000-0002-8197-6181}
\and
\IEEEauthorblockN{Qiancheng Fu}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
qcfu@bu.edu}
\and
\IEEEauthorblockN{Hang Ji\textsuperscript{\textsection}}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Cornell University}\\
Ithaca, USA \\
hj476@cornell.edu}
\and
\IEEEauthorblockN{Ines Santacruz}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
email address or ORCID}
\and
\IEEEauthorblockN{Marco Gaboardi}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Boston University}\\
Boston, USA \\
email address or ORCID}
}

@@

Guarded Kleene Algebra with Tests, or GKAT~\cite{smolka_GuardedKleeneAlgebra_2020}, is a deterministic fragment of Kleene Algebra with Tests. 
The syntax of GKAT over a set of primitive actions \(K\) and a set of primitive tests \(T\) can be defined in two sorts, boolean expressions \(\BExp\) and GKAT expressions \(\Exp\):
\begin{align*}
    a, b, c ∈ \BExp 
        & ≜ 1 ∣ 0 ∣ t ∈ T ∣ b ∧ c ∣ b ∨ c ∣ \overline{b} \\  
    e, f ∈ \Exp 
        & ≜ p ∈ K ∣ b ∈ \BExp ∣ e +_b f ∣ e ; f ∣ e^{(b)} 
\end{align*}
where \(e +_b f\) is the if statement with condition \(b\), \(e;f\) is the sequencing of expression \(e\) and \(f\), and \(e^{(b)}\) is the while loop with body \(e\) and condition \(b\).
We use notation like \(b ≤ c\), \(b ≡ c\), and \(b ≢ c\) for the usual order, equivalence, and inequivalence in Boolean Algebra.
A GKAT expression can be unfolded into a KAT expression in the usual manner~\cite{kozen_KleeneAlgebraTests_1997c}:
\begin{align*}
    e +_b f & ≜ b; e + \overline{b}; f &
    e^{(b)} & ≜ (b; e)^*; \overline{b}.
\end{align*}
Then the semantics of each expression \(⟦e⟧\) can be computed by the semantics of Kleene Algebra with Tests~\cite{kozen_KleeneAlgebraTests_1997c}.
An important construct in the semantics is \emph{atoms}, which are conjunctions of all the primitive tests either in its positive or negative form: for \(T ≜ \{t₁, t₂, …, tₙ\}\)
\[\At_T ≜ \{t₁' ∧ t₂' ∧ ⋯ ∧ tₙ' ∣ tᵢ' ∈ \{tᵢ, \overline{tᵢ}\}\}.\]
Follow the conventional notation, we denote an atom using \(α, β\); and we sometimes omit the subscript \(T\) when no confusing can arise.
Alternatively, atoms can also be thought of as truth assignments to each primitive tests, indicating which primitive tests is satisfied in the current program states; and \(α ≤ b\) if and only if the truth assignment represented by \(α\) satisfies \(b\).

For the sake of brevity, we omit the complete definition of GKAT and KAT semantics, we refer the reader to previous works~\cite{smolka_GuardedKleeneAlgebra_2020,schmid_GuardedKleeneAlgebra_2021,kozen_KleeneAlgebraTests_1997c}, which explain these semantics in detail.
Our work avoids direct interaction with the semantics by leveraging prior results in the coalgebraic theory of GKAT, which we will recap below.

@@

Besides nice property with bisimulation, GKAT coalgebra is also deterministic, unlike Kleene Coalgebra with Tests (KCT)~\cite{kozen_CoalgebraicTheoryKleene_2017}.
For each atom, states in a KCT can accept or reject the atom (but not both), yet states can also non-deterministically transition to multiple different states via the same atom, while executing different actions.
As we will see later, the deterministic behavior of GKAT coalgebra not only enables a more versatile symbolic algorithm than KCT~\cite{pous_SymbolicAlgorithmsLanguage_2015}, but also present challenges. 
Specifically, GKAT coalgebra requires normalization to compute finite trace equivalences~\cite{smolka_GuardedKleeneAlgebra_2020}, where we reroute all the transitions that cannot lead to acceptance immediately into rejection; and states that can never lead to acceptance are called \emph{dead states}.

In previous works, these dead states are detected after the necessary coalgebra is computed and stored.
This approach requires storing all the transitions and states of coalgebra in memory, meaning that the algorithm does not short circuit even if a mismatch can be detected in the starting state; for example, when for some \(α ∈ \At\), \(δ_S(s, α) = \reject\) and \(δ_U(u, α) = \accept\).

Our algorithm is lazy in the dead states detection i.e. the dead states are only checked when a mismatch requires it; even then, our dead state detection algorithm will only check the necessary states to compute the liveness of the states causing the mismatch.
This not only avoids unnecessarily checking the liveness of a state, but also enables on-the-fly generation of the coalgebra, like using derivatives, where we can remove irrelevant states from memory after they have been explored.

However, to truly understand our on-the-fly algorithms, we will first need to define ``dead states'', and their roles in defining the coalgebraic semantic of GKAT.

@@

Traditionally, live and dead states are defined by whether they can reach to an accepting state~\cite{smolka_GuardedKleeneAlgebra_2020}.
\begin{definition}[Liveness]\label{def:liveness-of-states}
    A state \(s\) is \emph{accepting} if there exists an atom \(α ∈ \At\) s.t. \(δ(s, α) = \accept\).
    A state \(s₀\) is \emph{live} if there exists an accepting state \(sₙ\) s.t. there is a liveness chain:
    \[s₀ \transvia{α₁ ∣ p₁} s₁ \transvia{α₂ ∣ p₂} s₂ ⋯ \transvia{αₙ ∣ pₙ} sₙ.\]
    A state is \emph{dead} if it is not dead.
\end{definition}
However, besides directly working with this explicit definition, we can also give a more coalgebraic characterization of these properties.
Specifically, by induction on the length of the path from one state to the next, we can show that the principle sub-coalgebra \(⟨s⟩_S\) exactly contains all the reachable state of \(s\) in the GKAT coalgebra \(S\).
Thus we obtain the following coalgebraic characterization of liveness:
\begin{theoremEnd}{theorem}\label{thm:liveness-principle-sub-coalg}
    A state \(s\) is \emph{live} if there exists an accepting state \(s' ∈ ⟨s⟩\); and a state \(s\) is \emph{dead} if there is no accepting state in \(⟨s⟩\).
\end{theoremEnd}
This alternative liveness definition can help us prove important theorems regarding reachability and liveness without explicitly performing induction on the liveness chain. 
We show the following theorems as examples:
\begin{theoremEnd}{lemma}\label{thm:dead-iff-all-reachable-dead}
    A state \(s\) is dead if and only if all elements in \(⟨s⟩\) is dead.
\end{theoremEnd}
\begin{proofEnd}
    \(⟸\) direction is true, because \(s ∈ ⟨s⟩\): if all \(⟨s⟩\) is dead, then \(s\) is dead. 
    \(⟹\) direction can be proven as follows.
    Take any \(s' ∈ ⟨s⟩\), then \(⟨s'⟩ ⊑ ⟨s⟩\) because \(s'\) is the minimal sub-coalgebra that contains \(s'\). 
    Since there is no accepting state in \(⟨s⟩\), thus there cannot be any accepting state in \(⟨s'⟩\), hence \(s'\) is also dead.
\end{proofEnd}

@@

Besides giving us the finite-trace semantics, the normalization operation also satisfies several nice properties.

@@

Second, normalization is an endofunctor in the category of GKAT coalgebra. This result can help us obtain sub-coalgebras of normalized coalgebra, and also connect the finite trace with infinite trace semantics.
